\chapter*{Acknowledgments}
Doing a PhD is not trivial (no citation needed here, I think). It is made possible not just from one's own hard work, but with help and support from many others. 

To start, the PhD was possible due to financial support from the German Academic Exchange Foundation (DAAD, in german). I am very thankful for the opportunity to do all this work with their warm support. In this sense, I am also grateful to the University of Oldenburg for the structures provided that allowed me to work. 

Moving from the material to the intellectual and personal support, I am extremely grateful to Prof. Ulrike Feudel, my supervisor for all of these years! We had a lot of fun, and I learned so much from her. On top of the specific knowledge of the area, I think maybe the most important skill she developed on me was the ability to refine a thought and rigorously develop an idea until you are confident about it. On top of all the actual science we made, I find it amazing how many conferences I had the opportunity to participate in, and how many people I met - a lot of this I owe to Ulrike's support. In a similar vein I also want to deeply thank my friend and collaborator, Prof. (yes, Prof.!) Everton Medeiros. Maybe without even noticing, he also taught me so much. He was a crucial aid in making this whole process a lot more fun and exciting - with, of course, better science. I will always remember this fondly - as will I remember his legendary goal from the corner of the pitch in Spiekeroog. Football's loss is academia's gain.       

Moving more to the emotional (and personal still) support, I am above all thankful to my mom and my sister for just about everything. Moving abroad is not easy - for anyone involved - but their unwavering love, warmth, and understanding made it all happen! This extends to my whole family, in fact. 

Friends are also a crucial ingredient to a nice life. And in this I also have had the luck to count on some amazing figures. I will only name some - their egos will have to be satisfied. In no particular order I want to thank from the bottom of my heart Jakob and Bianca Weik, Roberto Budzinski, Bruno Boaretto, Maira Theisen, Gabriel Gubert, Lucas Pollyceno, Carlos Martins, Alexandre Camargo, Rafa Jakuboski, Joao Paludo, Marcos Sato, George Datseris, Bryony Hobden, Patryk Bielski, Luis Gustavo, and Deoclecio Valente - and many others I equally love but, come on, the list was already too big. 

Finally, I mentioned I met a lot of amazing people doing science, some of whom I already mentioned. For all the discussions, support, and fun, I want to especially thank Klaus Lehnertz, Lyle Muller, Pete Ashwin, Aneta Koseska, Vyacheslav Kruglov, Ryan Deeley, and Jan Freund.