
\section*{List of Publications}
\addcontentsline{toc}{section}{List of Publications}  
This dissertation is based on the following publications:


\paragraph{Chapter 2.1.7:}
George Datseris, \underline{Kalel L. Rossi}, and Alexandre Wagemakers. Framework for global stability analysis of dynamical systems. \textit{Chaos} 33, 073151 (2023).

\paragraph{Chapter 3:}
\underline{Kalel L. Rossi}, Roberto C. Budzinski, Bruno R. R. Boaretto, Lyle E. Muller, and Ulrike Feudel.  Small changes at single nodes can shift global network dynamics. \textit{Physical Review Research} 5, 013220 (2023). 

\paragraph{Chapter 4:}
\underline{Kalel L. Rossi}, Everton S. Medeiros, Peter Ashwin and Ulrike Feudel. Transients versus network interactions.

\paragraph{Chapter 5:}
\underline{Kalel L. Rossi}, Roberto C. Budzinski, Bruno R. R. Boaretto, Lyle E. Muller, and Ulrike Feudel. Dynamical properties and mechanisms of metastability: a perspective in neuroscience. 

\vspace{1cm}


On top of these main thesis papers, I have also collaborated in other works, which resulted in three further publications, with me as a co-author.
\begin{itemize}
    % \item George Datseris, \underline{Kalel L. Rossi}, and Alexandre Wagemakers. Framework for global stability analysis of dynamical systems. \textit{Chaos} 33, 073151 (2023).
    \item Bruno R. R. Boaretto, Roberto C. Budzinski, \underline{Kalel L. Rossi}, Thiago L. Prado, Sergio R. Lopes and Cristina Masoller. Temporal Correlations in Time Series Using Permutation Entropy, Ordinal Probabilities and Machine Learning. \textit{Entropy} 23, 1025 (2021).
    \item Bruno R.R. Boaretto, Roberto C. Budzinski, \underline{Kalel L. Rossi}, Cristina Masoller, Elbert E.N. Macau. Spatial permutation entropy distinguishes resting brain states. \textit{Chaos, Solitons and Fractals} 171, 113453 (2023).
\end{itemize}